An empirical assessment of strategies to model party interactions in crash severity analysis

### 2020-05-10 Interactions-Crash-Severity-v1/Interactions-Crash-Severity-v2.tex
First round of edits to respond to reviewers after decision on original submission.

  \journal{Some Journal} % Sets Journal name

@@

% Pandoc header
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Evidence from across the world suggests that the burden of road
accidents is not borne evenly. There are important disparities at the
international level, where the odds of death due to road crashes are
three times higher in low-income countries compared to high-income
countries (World Health Organization, 2019). In the case of high-income
countries, where substantial gains in road safety have been observed for
years, said gains have also been unevenly distributed; thus, while fatal
crashes involving older adults in the United States and Great Britain
declined between 1997 and 2010 (despite the graying of the population),
the trend remained stable or increased slightly in Australia in roughly
the same period (Thompson et al., 2018). The impact of road accidents
also varies considerably. For example, in a study in the United States,
Obeng (2011) reported that the factors affecting crash severity varied
substantially in magnitude by gender. More recently, Regev et al.~(2018)
used adjusted crash risk to find that the risk of fatal injuries for
older drivers remained high even for minor collisions - which highlights
the perils of accidents at older ages. Other studies have concentrated
on the consequences of road accidents for the young (e.g., Peek-Asa et
al., 2010), the old (e.g., Rakotonirainy et al., 2012), as well as
pedestrians and cyclists (e.g., Hanson et al., 2013; McArthur et al.,
2014).

Given the relevance and cost of this matter, as well as the unevenness
of impacts among different population segments, numerous efforts have
been conducted to better understand the factors that affect road safety
- including the probable consequences of crashes. Along these lines, a
stream of research in the analysis of road accidents is concerned with
the severity of crashes. In particular, multivariate analysis of crash
severity is a useful way to clarify the way various factors can affect
the outcome of an incident, and to discriminate between various levels
of injury. This is an active area of research, and one where
methodological developments have aimed at improving the reliability,
accuracy, and precision of models (e.g., Savolainen et al., 2011; Bogue
et al., 2017; Shamsunnahar and Eluru, 2013). Of interest in the
literature on crash severity is the way different parties in a crash
interact to influence the severity of individual outcomes. The
importance of these interactions has been recognized in the existing
literature (e.g., Chiou et al., 2013; Lee and Li, 2014; Li et al., 2017;
Torrao et al., 2014), and a number of different modelling strategies
have been proposed. These include the use of \emph{data subsetting},
\emph{opponent variables}, and estimating models with \emph{hierarchical
components}. A gap in the literature, however, is a systematic
assessment of the performance of these various strategies.

The objective of this paper is to empirically evaluate different
strategies to model opponent effects in crash severity, in the context
of incidents involving two parties. Therefore, we present a systematic
assessment of several relevant modelling strategies. For the assessment
we use data from Canada's National Collision Database, a database that
collects all police-reported collisions in the country. Using the most
recent version of the data set (2017). Three levels of crash severity
(no injury/injury/fatality) are analyzed using ordered logit models, and
covariates for the participants in the crash and the conditions of the
crash. For model assessment, we conduct an in-sample prediction exercise
using the estimation sample (i.e., \emph{nowcasting}), and also an
out-of-sample prediction exercise using the data set corresponding to
2016 (i.e., \emph{backcasting}). The models are assessed using predicted
shares and predicted classes of outcomes at the individual level, using
an extensive array of verification statistics.

@@

Modelling the severity of injuries to participants in road crashes has
been a preoccupation of transportation researchers, planners, auto
insurance companies, governments, and the general public for decades.
One of the earliest studies to investigate the severity of injuries
conditional on an accident having occurred was by White and Clayton
(1972). Kim et al.~(1995) later stated that the ``linkages between
severity of injury and driver characteristics and behaviors have not
been thoroughly investigated'' (p.~470). Nowadays, there is a burgeoning
literature on this subject, including methodological developments, case
studies, and more niche research with a focus on particular situations
(e.g., crashes at intersections, Mussone et al., 2017; crashes in rural
roads, Gong and Fan, 2017), and crashes involving special population
groups (e.g., crashes involving motorcyclists or active travelers; see
Shaheed et al., 2013; Salon and McIntyre, 2018).

Crash severity is often modeled using models for discrete outcomes. An
analyst interested in crash severity has at their disposal an ample
repertoire of models to choose from, including classification techniques
from machine learning (e.g., Iranitalab and Khattak, 2017; Chang and
Wang, 2006; Effati et al., 2015; Khan et al., 2015), Poisson models for
counts (e.g., Ma et al., 2008), unordered logit/probit models (e.g., Tay
et al., 2011), as well as ordered logit/probit models (e.g., Rifaat and
Chin, 2007), with numerous variants, such as random parameters/mixed
logit (e.g., Aziz et al., 2013; Haleem and Gan, 2013), partial
proportional odds models (e.g., Mooradian et al., 2013; Sasidharan and
Menendez, 2014), and the use of copulas (e.g., Wang et al., 2015).
Recent reviews of methods include Savolainen et al.~(2011), Shamsunnahar
and Eluru (2013), and Mannering et al.~(2016).

Irrespective of the modelling framework employed, models of crash
severity often include variables in several categories, as shown with
examples in Table \ref{tab:variable-categories} (also see Montella et
al., 2013). Many crash databases and analyses also account for the
multievent nature of many crashes. Participants may have had different
roles in a crash depending on their context, with some acting as
operators of a vehicle (i.e., drivers, bicyclists), while others were
passengers. They also may differ depending on what type of traffic unit
they were, for example pedestrians, or operators/passengers of a
vehicle. The multiplicity of roles makes for complicated modelling
decisions when trying to understand the severity of injuries; for
example, what is the unit of analysis, the person, the traffic unit, or
the collision? Not surprisingly, it is possible to find examples of
studies that adopt different perspectives. A common simplifying strategy
in model specification is to consider only \emph{drivers} and/or only
\emph{single-vehicle} crashes (e.g., Kim et al., 2013; Gong and Fan,
2017; Lee and Li, 2014; Osman et al., 2018). This strategy reduces the
dimensions of the event, and it becomes possible, for example, to equate
the traffic unit to the person for modelling purposes.

The situation becomes more complex when dealing with events that involve
two traffic units (e.g., Torrao et al., 2014; Wang et al., 2015) and
multi-traffic unit crashes (e.g., Wu et al., 2014; Bogue et al., 2017).
Different strategies have been developed to study these, more complex
events. A number of studies advocate the estimation of separate models
for different participants and/or situations. In this way, Wang and
Kockelman (2005) estimated models for single-vehicle and two-vehicle
crashes, while Savolainen and Mannering (2007) estimated models for
single-vehicle and multi-vehicle crashes. More recently, Duddu et
al.~(2018) and Penmetsa et al.~(2017) presented research that estimated
separate models for at-fault and not-at-fault drivers. The strategy of
estimating separate models relies on \emph{subsetting} the data set,
although it is possible to link the relevant models more tightly by
means of a common covariance structure, as is the case of bivariate
models (e.g., Chiou et al., 2013; Chen et al., 2019) or models with
copulas (e.g., Rana et al., 2010; Shamsunnahar et al., 2014; Wang et
al., 2015).

A related strategy to specify crash severity models is to organize the
data in such a way that it is possible to introduce \emph{opponent
effects}. There are numerous examples of studies that consider at least
some characteristics of the opponents in two- or multi-vehicle crashes.
For example, Wang and Kockelman (2005) considered the type of the
opposing vehicle in their model for two-vehicle collisions. Similarly,
Torrao et al.~(2014) included in their analysis the age, wheelbase,
weight, and engine size of the opposing vehicle, while Bogue et
al.~(2017) used the body type of the opposing vehicle. Penmetsa et
al.~(2017) and Duddu et al.~(2018) are two of the most comprehensive
examples of using opponent's information, as they used individual
attributes of opponents (their physical condition, sex, and age), as
well as characteristics of the opposing traffic unit (the vehicle type
of the opponent). The twin strategies of subsetting the sample and using
the attributes of the opponent are not mutually exclusive, but neither
are they used consistently together, as a scan of the literature
reveals. Another strategy is to introduce \emph{hierarchical components}
in the model, a technique widely used in the hierarchical or multi-level
modelling literature.

@@

\begin{itemize}
\tightlist
\item
  Strategy 1. Introducing opponent-related factors
\item
  Strategy 2. Single-level model and multi-level (hierarchical) model
  specifications
\item
  Strategy 3. Full sample and sample subsetting
\end{itemize}

@@

\caption{\label{tab:table-variable-categories}\label{tab:variable-categories}Categories of variables used in the analysis of crash severity with examples}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{22em}}
\toprule
Category & Examples\\
\midrule
Human-factors & Attributes of participants in the crash, e.g., injury status, age, gender, licensing status, professional driver status\\
Traffic unit-factors & Attributes of the traffic unit, e.g., type of traffic unit (pedestrian, car, motorcycle, etc.), maneuver, etc.\\
Environmental-factors & Attributes of the crash, e.g., location, weather conditions, light conditions, number of parties, etc.\\
Road-factors & Attributes of the road, e.g., surface condition, grade, geometry, etc.\\
Opponent-related factors & Attributes of the opponent, e.g., age of opponent, gender of opponent, opponent vehicle type, etc.\\
\bottomrule
\end{tabular}
\end{table}

@@

Shamsunnahar and Eluru (2013) conducted an extensive comparison of
models for discrete outcomes in crash severity analysis, and found only
small differences in the performance of unordered models and ordered
models; however, ordered models are usually more parsimonious since only
one latent functions needs to be estimated for all outcomes, as opposed
to one for each outcome in unordered models. Bogue et al.~(2017) also
compared unordered and ordered models in the form of the mixed
multinomial logit and a modified rank ordered logit, respectively, and
found that the ordered model performed best. To keep the empirical
assessment manageable, in this paper we will consider only the ordinal
logit model, and will comment on potential extensions in Section
\ref{sec:further-considerations}.

@@

The left-hand side of the expression above (\(y_{itk}^*\)) is a latent
(unobservable) variable that is associated with the severity of crash
\(k\) (\(k=1,\cdots,K\)) for participant \(i\) in traffic unit \(t\).
The right-hand side of the expression is split in four parts. The first
part collects \(l=1,\cdots,L\) human-factors \(p\) for participant \(i\)
in traffic unit \(t\) and crash \(k\); these could relate to the person
(e.g., age, gender, and road user class). The second part collects
\(m=1,\cdots,M\) attributes \(u\) related to traffic unit \(t\) in crash
\(k\); these could be items such as maneuver or vehicle type. The third
part collects \(q=1,\cdots,Q\) attributes \(c\) related to the crash
\(k\), including environmental-factors and road-factors, such as weather
conditions, road alignment, and type of surface. Lastly, the fourth
element is a random term specific to participant \(i\) in traffic unit
\(t\) and crash \(k\). The function consists of a total of \(Z=L+M+Q\)
covariates and associated parameters.

@@

The latent variable is not observed directly, but it is possible to
posit a probabilistic relationship with the outcome \(y_{itk}\) (the
severity of crash \(k\) for participant \(i\) in traffic unit \(t\)).
Depending on the characteristics of the data and the assumptions made
about the random component of the latent function different models can
be obtained. For example, if crash severity is coded as a variable with
three outcomes (e.g., property damage only/injury/fatal), we can relate
the latent variable to the outcome as follows:

@@

The additional part collects \(r=1,\cdots,R\) attributes \(o\) related
to individual \(j\) in traffic unit \(v\) and crash \(k\) that opposed
individual \(i\) in traffic unit \(t\) and crash \(k\). These could be
individual characteristics of the opponent (such as age and gender)
and/or characteristics of the opposing vehicle (such as vehicle type or
weight). To qualify as an opponent, individual \(j\) must have been a
participant in crash \(k\) but operating traffic unit \(v\ne t\).
Sometimes the person \emph{is} the traffic unit, as is the case of a
pedestrian. And we exclude passengers of vehicles as opponents, since
they do not operate the traffic unit. In case the opponent attributes
include only characteristics of the traffic unit, the condition for the
traffic unit to be an opponent is that it participated in crash \(k\)
and was different from \(t\). After introducing this new set of terms,
the latent function now consists of a total of \(Z=L+M+Q+R\) covariates
and associated parameters.

@@

The coefficients of the traffic unit nest the individual attributes as
follows. For any given coefficient \(q\):

@@

\begin{quote}
\textbf{Note:} this paper presents reproducible research. The source
file is an R Markdown document. All code and data necessary to reproduce
the analysis are available from the following anonymous Drive folder:
\end{quote}

\begin{quote}
https://drive.google.com/open?id=12aJtVBaQ4Zj0xa7mtfqxh0E48hKCb\_XV
\end{quote}

\begin{quote}
The source files, code, and data will be publicly available in a GitHub
repository upon acceptance of the paper for publication
\end{quote}

@@

The NCDB is available from 1999. For the purpose of this paper, we use
the data corresponding to 2017, which is the most recent year available
as of this writing. Furthermore, for assessment we also use the data
corresponding to 2016. Similar to databases in other jurisdictions (see
Montella et al., 2013), the NCDB contains information pertaining to the
collision, the traffic unit(s), and the person(s) involved in a crash.
The definitions of variables in this database are shown in the Appendix
at the end of this document, in Tables
\ref{tab:ncdb-descriptives-collision},
\ref{tab:ncdb-descriptives-vehicle}, and
\ref{tab:ncdb-descriptives-person}. Notice that, compared to Table
\ref{tab:variable-categories}, environmental-factors variables and
road-factors variables are collected under a single variable class,
namely collision-related, since they are unique for each crash.

Data are organized by person; in other words, there is one record per
participant in a collision, be they drivers, passengers, pedestrians,
etc. The only variable directly available with respect to opponents in a
collision is the number of vehicles involved (see models in Bogue et
al., 2017). Therefore, the data needs to be processed to obtain
attributes of opponents for each participant in a collision. The
protocol to do this is described next.

\hypertarget{data-preprocessing-initial-filter}{%
\subsubsection{Data preprocessing: initial
filter}\label{data-preprocessing-initial-filter}}

We apply an initial filter, whereby we scan the database to remove all
records that are not a person (including parked cars and other objects)
or that are missing information (as is the case when a participant in a
crash is a runaway vehicle). Next, records missing at least one of the
next variables are removed: P\_USER (the road user class), P\_SEX (sex),
P\_AGE (age), and P\_ISEV (individual-level severity of crash). This
initial filter ensures that all records are complete from the
perspective of key information for analysis.

\hypertarget{data-preprocessing-filter-two-vehicle-collisions}{%
\subsubsection{Data preprocessing: filter two-vehicle
collisions}\label{data-preprocessing-filter-two-vehicle-collisions}}

After the initial filter, the database is summarized to find the number
of individual-level records that correspond to each collision (C\_CASE).
At this point, there are 32,298 collisions, involving only one (known)
participant, there are 46,483 collisions involving two participants,
19,433 collisions with three participants, 8,250 collisions involving
four participants, 3,783 collisions with five participants, 1,789
collisions with six participants, and 1,491 collisions involving seven
or more participants. These participants were possibly occupants in
different vehicles or were otherwise their own traffic units.
Accordingly, the sample includes 174,741 drivers, 61,403 passengers,
10,798 pedestrians, 5,286 bicyclists, and 6,564 motorcyclists.

The next step in our data preprocessing protocol is to remove all
collisions that involve only one participant. This still leaves numerous
cases where multiple participants could have been in a single vehicle,
for instance in a collision against a stationary object. Therefore, we
proceed to use the vehicle sequence number to find the number of
vehicles involved in each collision. This reveals that there are 20,732
collisions that involve only one vehicle but possibly multiple
participants (i.e., driver and one or more passengers). In addition,
there are 165,520 collisions involving two vehicles (and possibly
multiple participants). Finally, there are 40,242 collisions with three
or more vehicles.

Once we have identified the number of vehicles in each collision, we
proceed to select all cases that involve only two vehicles. The most
common cases in two-vehicle collisions are those that include drivers
(40,297 collisions; this is reflective of the prevalence of
single-occupant vehicles). This is followed by cases with passengers
(14,120 collisions), pedestrians (5,204 collisions), bicyclists (2,238
collisions), and motorcyclists (1,016 collisions). The distribution of
individuals per traffic unit is as follows: 80,382 individuals are coded
as being in V\_ID = 1, 76,523 individuals are coded as being in V\_ID =
2, and 7,932 individuals are coded as pedestrians. In addition, 683
individuals are coded as being in vehicles 3 through 9, despite our
earlier filter to retain only collisions with two vehicles. We therefore
proceed to select only individuals assigned to vehicles 1 or 2, as well
as pedestrians. As a result of this filter a number of cases with only
one known participant need to be removed.

\hypertarget{data-preprocessing-extract-opponent-information-and-join-to-individual-records}{%
\subsubsection{Data preprocessing: extract opponent information and join
to individual
records}\label{data-preprocessing-extract-opponent-information-and-join-to-individual-records}}

Up to this point, the goal of data preprocessing has been to obtain a
complete, workable sample of individual records of participants in
two-vehicle collisions. There are two possible cases for the collisions,
depending on the traffic units involved: 1) vehicle vs vehicle
collisions (``vehicle'' is all motorized vehicles, including
motorcycles/mopeds, as well as bicycles); and 2) vehicle vs pedestrian
collisions. To identify opponents in a collision, it is convenient to
classify collisions by pedestrian involvement. In this way, we find that
the database includes 16,636 collisions that are vehicle vs pedestrian
(possibly multiple pedestrians), and 147,594 collisions that involve two
vehicles. After splitting the database according to pedestrian
involvement, we can now extract relevant information about participants
in the collision. This involves renaming the person-level variables so
that we can distinguish each individual by their role as an individual
or an opponent in a given record. Notice that when working with
individuals in vehicles, only the driver is considered a legitimate
opponent in a collision (i.e., passengers are never considered
opponents).

Once the personal attributes of individuals that can be considered
opponents in a given collision have been extracted, their information is
joined to the individual records by means of the collision unique
identifier. As a result of this process, a new set of variables are now
available for analysis: the age, sex, and road user class of the
opponent, as well as the type of the opposing vehicle. A summary of
opponent interactions and outcomes can be found in Table
\ref{tab:summary-opponent-information}. The information there shows that
the most common type of opponent for drivers is other drivers, followed
by pedestrians. The only opponents of pedestrians, on the other hand,
are drivers. Bicyclists and motorcyclists, on the other hand, are mostly
opposed by drivers, but occasionally by other road users as well. In
terms of outcomes, we observe that virtually all fatalities occur when
the opponent is a driver, and only very rarely when the opponent is a
motorcyclist. Injuries are also more common when the opponent is a
driver, whereas no injuries are relatively more frequent when the
opponent is a pedestrian or a bicyclist.

@@

>>>>

  \journal{Accident Analysis and Prevention} % Sets Journal name

@@

% Pandoc header
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\doublespacing
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Modelling the severity of injuries to victims of road crashes has been a
preoccupation of transportation researchers, planners, auto insurance
companies, governments, and the general public for decades. One of the
earliest studies to investigate the severity of injuries conditional on
a collision having occurred was by White and Clayton (1972). Kim et
al.~(1995) later stated that the ``linkages between severity of injury
and driver characteristics and behaviors have not been thoroughly
investigated'' (p.~470). Nowadays, there is a burgeoning literature on
this subject, which often relies on multivariate analysis of crash
severity to clarify the way various factors can affect the outcome of an
incident, and to discriminate between various levels of injury.

Crash severity is an active area of research, and one where
methodological developments have aimed at improving the reliability,
accuracy, and precision of models (e.g., Savolainen et al., 2011; Bogue
et al., 2017; Yasmin and Eluru, 2013). Of interest in this literature is
how different parties in a crash interact to influence the severity of
individual outcomes. The importance of these interactions has been
recognized by numerous authors (e.g., Chiou et al., 2013; Lee and Li,
2014; Li et al., 2017; Torrao et al., 2014). Lee and Li (2014) for
instance, note that ``{[}for{]} two-vehicle crashes, most studies only
considered the effects of one vehicle on driver's injury severity or the
highest injury severity of two drivers. However, it is expected that
driver's injury severity is not only affected by characteristics of
his/her own vehicle, but also characteristics of a partner vehicle.''
More generally, the severity of the outcome depends, at least in part,
on the characteristics of the parties, and a crash between two heavy
vehicles is likely to have very different outcomes compared to crash
where a heavy vehicle hits a pedestrian.

For the purpose of this paper, we define a party as one or more
individuals travelling in a traffic unit that becomes involved in a
crash. Sometimes the traffic unit is a vehicle, and the party is a
single individual (i.e.~a single occupant vehicle); but in other cases,
a party could consist of several individuals (i.e., a driver and one or
more passengers). Other times, the individual \emph{is} the traffic
unit, for instance a pedestrian or a bicyclist. An opponent is a
different party that is involved in the same collision. In the
literature, interactions between parties in a collision are treated by
means of different strategies, including the use of \emph{data
subsetting}, \emph{opponent variables}, and estimating models with
\emph{hierarchical components}. These strategies are not new, however, a
systematic comparison between them is missing from the literature. For
this reason, the objective of this paper is to empirically evaluate
different strategies to model party interactions in crash severity in
the context of incidents involving two parties. More concretely, this
research aims to:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Systematize the analysis of party interactions in crash severity
  models. Although every strategy considered here has been used in past
  research, in this paper they are organized in a way that clarifies
  their operation.
\item
  Present a data management workflow to prepare a dataset to implement
  analysis of opponent effects.
\item
  Provide evidence of the performance of different modelling strategies.
  In particular, the importance of considering opponent-level effects
  and the suitability of single-level models.
\item
  Present an example of reproducible research in crash severity
  analysis: all data and code are publicly available from the beginning
  of the peer-review
  process\footnote{The source file for this paper is an `R` Markdown document; all code and data necessary to reproduce the analysis are available from the following GitHub repository: \url{https://github.com/paezha/Modelling-Participant-Interactions-in-Crash-Severity}}.
\end{enumerate}

For the assessment we use data from Canada's National Collision
Database, a database that collects all police-reported collisions in the
country. Using the most recent version of the data set (2017). Three
levels of crash severity (no injury/injury/fatality) are analyzed using
ordered logit models, and covariates for the parties in the crash and
the conditions of the crash. For model assessment, we conduct an
in-sample prediction exercise using the estimation sample (i.e.,
\emph{nowcasting}), and also an out-of-sample prediction exercise using
the data set corresponding to 2016 (i.e., \emph{backcasting}). The
models are assessed using predicted shares and predicted classes of
outcomes at the individual level, using an extensive array of
verification statistics.

@@

Crash severity is often modeled using models for discrete outcomes.
Analysts interested in crash severity have at their disposal an ample
repertoire of models to choose from, including classification techniques
from machine learning (e.g., Iranitalab and Khattak, 2017; Chang and
Wang, 2006; Effati et al., 2015; Khan et al., 2015), Poisson models for
counts (e.g., Ma et al., 2008), unordered logit/probit models (e.g., Tay
et al., 2011), as well as ordered logit/probit models (e.g., Rifaat and
Chin, 2007), with numerous variants, such as random parameters/mixed
logit (e.g., Aziz et al., 2013; Haleem and Gan, 2013), partial
proportional odds models (e.g., Mooradian et al., 2013; Sasidharan and
Menendez, 2014), and the use of copulas (e.g., Wang et al., 2015).
Recent reviews of methods include Savolainen et al.~(2011), Yasmin and
Eluru (2013), and Mannering et al.~(2016).

Irrespective of the modelling framework employed, models of crash
severity often include variables in several categories, as shown with
examples in Table \ref{tab:variable-categories} (also see Montella et
al., 2013). Many crash databases and analyses also account for the
multi-event nature of many crashes. Individuals in the crash may have
had different roles depending on their situation, with some acting as
operators of a vehicle (i.e., drivers, bicyclists), while others were
passengers. They also may differ depending on what type of traffic unit
they were, for example pedestrians, or operators/passengers of a
vehicle. The multiplicity of roles makes for complicated modelling
decisions when trying to understand the severity of injuries; for
example, what is the unit of analysis, the person, the traffic unit, or
the collision? Not surprisingly, it is possible to find examples of
studies that adopt different perspectives. A common simplifying strategy
in model specification is to consider only \emph{drivers} and/or only
\emph{single-vehicle} crashes (e.g., Kim et al., 2013; Gong and Fan,
2017; Lee and Li, 2014; Osman et al., 2018). This strategy reduces the
dimensions of the event, and it becomes possible, for example, to equate
the traffic unit to the person for modelling purposes.

The situation becomes more complex when dealing with events that involve
two traffic units (e.g., Torrao et al., 2014; Wang et al., 2015) and
multi-traffic unit crashes (e.g., Wu et al., 2014; Bogue et al., 2017).
Different strategies have been developed to study these, more complex
events. A number of studies advocate the estimation of separate models
for different parties, individuals, and/or situations. In this way, Wang
and Kockelman (2005) estimated models for single-vehicle and two-vehicle
crashes, while Savolainen and Mannering (2007) estimated models for
single-vehicle and multi-vehicle crashes. More recently, Duddu et
al.~(2018) and Penmetsa et al.~(2017) presented research that estimated
separate models for at-fault and not-at-fault drivers. The strategy of
estimating separate models relies on \emph{subsetting} the data set,
although it is possible to link the relevant models more tightly by
means of a common covariance structure, as is the case of bivariate
models (e.g., Chiou et al., 2013; Chen et al., 2019) or models with
copulas (e.g., Rana et al., 2010; Shamsunnahar et al., 2014; Wang et
al., 2015).

A related strategy to specify crash severity models is to organize the
data in such a way that it is possible to introduce \emph{opponent
effects}. There are numerous examples of studies that consider at least
some characteristics of the opposite party (or parties) in two- or
multi-vehicle crashes. For example, Wang and Kockelman (2005) considered
the type of the opposite vehicle in their model for two-vehicle
collisions. Similarly, Torrao et al.~(2014) included in their analysis
the age, wheelbase, weight, and engine size of the opposite vehicle,
while Bogue et al.~(2017) used the body type of the opposite vehicle.
Penmetsa et al.~(2017) and Duddu et al.~(2018) are two of the most
comprehensive examples of using opponent's information, as they included
attributes of individuals in the opposite party (their physical
condition, sex, and age), as well as characteristics of the other
party's traffic unit (the vehicle type of the opponent). The twin
strategies of subsetting the sample and using the attributes of the
opponent are not mutually exclusive, but neither are they used
consistently together, as a scan of the literature reveals.

Another strategy is to introduce \emph{hierarchical components} in the
model, a technique widely used in the hierarchical or multi-level
modelling literature. This involves considering observations as being
nested at different levels: individuals nested in traffic units, which
in turn are nested in accidents, as an example.

@@

\begin{itemize}
\item
  Strategy 1. Introducing opponent-related factors
\item
  Strategy 2. Single-level model and multi-level (hierarchical) model
  specifications
\item
  Strategy 3. Full sample and sample subsetting
\end{itemize}

@@

\caption{\label{tab:table-variable-categories}\label{tab:variable-categories}Categories of variables used in the analysis of crash severity with examples}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{22em}}
\toprule
Category & Examples\\
\midrule
Human-factors & Attributes of individuals in the crash, e.g., injury status, age, gender, licensing status, professional driver status\\
Traffic unit-factors & Attributes of the traffic unit, e.g., type of traffic unit (pedestrian, car, motorcycle, etc.), maneuver, etc.\\
Environmental-factors & Attributes of the crash, e.g., location, weather conditions, light conditions, number of parties, etc.\\
Road-factors & Attributes of the road, e.g., surface condition, grade, geometry, etc.\\
Opponent-related factors & Attributes of the opponent, e.g., age of opponent, gender of opponent, opponent vehicle type, etc.\\
\bottomrule
\end{tabular}
\end{table}

@@

Before describing the modelling strategies, it is important to explain
our choice of model. There have already been comparisons between
different models. Yasmin and Eluru (2013), for instance, conducted an
extensive comparison of models for discrete outcomes in crash severity
analysis, and found only small differences in the performance of
unordered models and ordered models; however, ordered models are usually
more parsimonious since only one latent functions needs to be estimated
for all outcomes, as opposed to one for each outcome in unordered
models. Bogue et al.~(2017) also compared unordered and ordered models
in the form of the mixed multinomial logit and a modified rank ordered
logit, respectively, and found that the ordered model performed best. To
keep the empirical assessment manageable, in this paper we will consider
only the ordinal logit model, and will comment on potential extensions
in Section \ref{sec:further-considerations}.

@@

The left-hand side of the expression above (\(y_{itk}^*\)) is a latent
(unobservable) variable that is associated with the severity of crash
\(k\) (\(k=1,\cdots,K\)) for individual \(i\) in traffic unit \(t\). The
right-hand side of the expression is split in four parts. The first part
gathers \(l=1,\cdots,L\) human-factors \(p\) for individual \(i\) in
traffic unit \(t\) and crash \(k\); these could relate to the person
(e.g., age, gender, and road user class). The second part gathers
\(m=1,\cdots,M\) attributes \(u\) related to traffic unit \(t\) in crash
\(k\); these could be items such as maneuver or vehicle type. The third
part gathers \(q=1,\cdots,Q\) attributes \(c\) related to the crash
\(k\), including environmental-factors and road-factors, such as weather
conditions, road alignment, and type of surface. Lastly, the fourth
element is a random term specific to individual \(i\) in traffic unit
\(t\) and crash \(k\). The function consists of a total of \(Z=L+M+Q\)
covariates and associated parameters.

@@

The latent variable is not observed directly, but it is possible to
posit a probabilistic relationship with the outcome \(y_{itk}\) (the
severity of crash \(k\) for individual \(i\) in traffic unit \(t\)).
Depending on the characteristics of the data and the assumptions made
about the random component of the latent function different models can
be obtained. For example, if crash severity is coded as a variable with
three outcomes (e.g., property damage only/injury/fatal), we can relate
the latent variable to the outcome as follows:

@@

This function includes one additional summation compared to Equation
\ref{eq:latent-function-ordered-outcomes}. This summation gathers
\(r=1,\cdots,R\) attributes \(o\) related to individual \(j\) in traffic
unit \(v\) that was an opposite party to individual \(i\) in traffic
unit \(t\) in crash \(k\). These attributes could be individual
characteristics of the operator of the opposite traffic unit (such as
age and gender) and/or characteristics of the opposite traffic unit
(such as vehicle type or weight). To qualify as an opposite party,
individual \(j\) must have been an individual in crash \(k\) but
operating traffic unit \(v\ne t\). Sometimes the person \emph{is} the
traffic unit, as is the case of a pedestrian. And we exclude passengers
of vehicles as opponents, since they do not operate the traffic unit. In
case the opponent attributes include only characteristics of the traffic
unit, the condition for the traffic unit to be an opponent is that it
was part of crash \(k\) and was different from \(t\). After introducing
this new set of terms, the latent function now consists of a total of
\(Z=L+M+Q+R\) covariates and associated parameters.

@@

The coefficients of the traffic unit nest the individual attributes as
follows. For any given coefficient \(m\):

@@

@@

The NCDB is available from 1999. For the purpose of this paper, we use
the data corresponding to 2017, which is the most recent year available
as of this writing. Furthermore, for assessment we also use the data
corresponding to 2016. Similar to databases in other jurisdictions (see
Montella et al., 2013), the NCDB contains information pertaining to the
collision, the traffic unit(s), and the person(s) involved in a crash.
The definitions of variables in this database are shown in the Appendix
at the end of this document, in Tables
\ref{tab:ncdb-descriptives-collision},
\ref{tab:ncdb-descriptives-vehicle}, and
\ref{tab:ncdb-descriptives-person}. Notice that, compared to Table
\ref{tab:variable-categories}, environmental-factors variables and
road-factors variables are gathered under a single variable class,
namely collision-related, since they are unique for each crash.

Data are organized by person; in other words, there is one record per
individual in a collision, be they drivers, passengers, pedestrians,
etc. The only variable directly available with respect to opponents in a
collision is the number of vehicles involved (see models in Bogue et
al., 2017). Therefore, the data needs to be processed to obtain
attributes of the opposing party for each individual in a collision. The
protocol to do this is described next.

\hypertarget{data-preprocessing}{%
\subsection{Data preprocessing}\label{data-preprocessing}}

To prepare the data for analysis, in particular for Strategy 1
(opponent-related factors), we apply an initial filter, whereby we scan
the database to remove all records that are not a person (including
parked cars and other objects) or that are missing information.

After the initial filter, the database is summarized to find the number
of individual-level records that correspond to each collision (C\_CASE).
At this point, there are 32,298 collisions, involving only one (known)
individual, there are 46,483 collisions involving two parties, 19,433
collisions with three parties, 8,250 collisions involving four parties,
3,783 collisions with five parties, 1,789 collisions with six parties,
and 1,491 collisions involving seven or more parties These parties were
possibly occupants in different vehicles or were otherwise their own
traffic units (e.g., pedestrians). Accordingly, the sample includes
174,741 drivers, 61,403 passengers, 10,798 pedestrians, 5,286
bicyclists, and 6,564 motorcyclists.

The next step is to remove all collisions that involve only one party.
This still leaves numerous cases where multiple parties could have been
in a single vehicle, for instance in a collision against a stationary
object. Therefore, we proceed to use the vehicle sequence number to find
the number of vehicles involved in each collision. This reveals that
there are 20,732 collisions that involve only one vehicle but possibly
multiple individuals (i.e., driver and one or more passengers). In
addition, there are 165,520 collisions involving two vehicles (and
possibly multiple individuals). Finally, there are 40,242 collisions
with three or more vehicles.

Once we have identified the number of vehicles in each collision, we
select all cases that involve only two vehicles. The most common cases
in two-vehicle collisions are those that include drivers (40,297
collisions; this is reflective of the prevalence of single-occupant
vehicles). This is followed by cases with passengers (14,120
collisions), pedestrians (5,204 collisions), bicyclists (2,238
collisions), and motorcyclists (1,016 collisions). The distribution of
individuals per traffic unit is as follows: 80,382 individuals are coded
as being in V\_ID = 1, 76,523 individuals are coded as being in V\_ID =
2, and 7,932 individuals are coded as pedestrians. In addition, 683
individuals are coded as being in vehicles 3 through 9, despite our
earlier filter to retain only collisions with two vehicles. At this
point we select only individuals assigned to vehicles 1 or 2, as well as
pedestrians. As a result of this filter a number of cases with only one
known individual need to be removed.

@@

At this point we have a complete, workable sample of individual records
of parties in two-vehicle collisions. There are two possible cases for
the collisions, depending on the traffic units involved: 1) vehicle vs
vehicle collisions (``vehicle'' is all motorized vehicles, including
motorcycles/mopeds, as well as bicycles); and 2) vehicle vs pedestrian
collisions. To identify the opposite parties in each collision it is
convenient to classify collisions by pedestrian involvement. In this
way, we find that the database includes 16,636 collisions that are
vehicle vs pedestrian (possibly multiple pedestrians), and 147,594
collisions that involve two vehicles. After splitting the database
according to pedestrian involvement, we can now extract relevant
information about the different parties in the collision. This involves
renaming the person-level variables so that we can distinguish each
individual by their party in a given record. Notice that when working
with individuals in vehicles, only drivers are considered opposites in a
collision.
